---
title: "Analysis of lipidomics data"
author: "Timothy Yu"
output: 
  github_document:
    toc: true
    html_preview: false
editor_options:
  chunk_output_type: inline
---
This notebook analyzes the liver and plasma lipidomics datasets.
```{r setup, include = FALSE}
library(tidyverse)
library(ggfortify)
library(bnstruct)
library(viridis)
library(DescTools)
library(RColorBrewer)
require(cowplot)
theme_set(theme_cowplot())
source('utils/utils.R')
```

```{r}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

if(!file.exists('../markdowns/lipidomics_files')){
  dir.create(file.path('../markdowns/lipidomics_files'))
  
}
```

Read in liver lipidomics dataset.
```{r}
liverData = read.csv('../processed_data/datasets/Lipidomics_liver_normliverweight.csv', header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
```

By 90% confidence ellipse, Mice 53 and 39 are outliers. We will remove them and re-plot the PCA. They are excluded from the remaining analysis.
```{r, echo=T, fig.width=3, fig.height=2, fig.align="center", dpi=300, dev="png"}
autoplot(prcomp((liverData[-c(1,2)] %>% select_if(~ !any(is.na(.)))), scale. = TRUE), 
         data = liverData, colour = 'Treatment', 
         size = 0.1, label = TRUE, label.size = 3, frame = TRUE, frame.type = 'norm', 
         frame.level = 0.90) + 
         scale_color_manual(values = c('black', '#3b9bb3', '#75af4f', '#e39225', '#c13b41')) +
         scale_fill_manual(values = c('black', '#3b9bb3', '#75af4f', '#e39225', '#c13b41'))
```

PCA plot of liver lipidomics data.
```{r, echo=T, fig.width=3, fig.height=2, fig.align="center", dpi=300, dev="png"}
liverData = liverData %>% filter(!ID %in% c(39,53))
autoplot(prcomp((liverData[-c(1,2)] %>% select_if(~ !any(is.na(.)))), scale. = TRUE), 
         data = liverData, colour = 'Treatment', size = 2.5) + 
  scale_color_manual(values = c('black', '#3b9bb3', '#75af4f', '#e39225', '#c13b41'))

#ggsave('../figures/lipidomics_liver_pca.pdf', height = 3.5, width = 5)
```
Bubble plot showing relative fold-change in liver lipid abundance.
```{r, echo=T, fig.width=7, fig.height=3, fig.align="center", dpi=300, dev="png"}
colors <- c("#A00000", "#00C000", "#5757F9", "#FF6000", "#0000C0", "#C0C000", "#D7B0B0",
            "#9F044D", "#077E97", "#C5944E", "#034E61", "#FFA040", "#606060", "#fcfc81")

bubdata = makeBubbleData(liverData)
bubp = getBubbleSignificance(bubdata)
makeBubblePlot(bubp, abundance=FALSE, colors=colors, bubble_range=c(3,10))

#ggsave('../figures/lipidomics_liver_bubble_plot.pdf', height = 5.5, width = 15)
#write.csv(bubp, '../processed_data/misc/lipidomics_liver_bubble_stats.csv')
```
Abundance of individual hepatic TG species split by SFA, MUFA, and PUFA.
```{r, echo=T, fig.width=4.5, fig.height=1.5, fig.align="center", dpi=300, dev="png"}
scaled_liverData = cbind(liverData[,c(1,2)], scale(liverData[,-c(1,2)])) # mean 0, sd 1

mean_TG_data = scaled_liverData %>%
  select_if(~ !any(is.na(.))) %>%
  gather(Feature, Value, 3:ncol(.)) %>%
  separate(col = 'Feature', into = c("Class", "Chain"), sep = "\\.", 
          remove = FALSE, extra = 'drop') %>%
  filter(Class == 'TG') %>%
  group_by(Treatment, Feature, Chain) %>%
  mutate(Mean_abundance = mean(Value)) %>%
  select(Treatment, Feature, Chain, Mean_abundance) %>%
  distinct() %>%
  mutate(Saturation = case_when(
    str_extract(Feature, "TG\\.\\d+\\.\\d+") %>% 
      str_extract("\\.\\d+$") %>%
      str_replace("\\.", "") %>%
      as.numeric() == 0 ~ "SFA",
    str_extract(Feature, "TG\\.\\d+\\.\\d+") %>% 
      str_extract("\\.\\d+$") %>%
      str_replace("\\.", "") %>%
      as.numeric() == 1 ~ "MUFA",
    str_extract(Feature, "TG\\.\\d+\\.\\d+") %>% 
      str_extract("\\.\\d+$") %>%
      str_replace("\\.", "") %>%
      as.numeric() > 1 ~ "PUFA",
    TRUE ~ NA_character_
  )) %>% mutate(Treatment = as.factor(Treatment)) %>% mutate(
    Treatment = fct_recode(
      Treatment,
      "0" = "Chow8",
      "2" = "WD2",
      "4" = "WD4",
      "6" = "WD6",
      "8" = "WD8"
  )
)

mean_TG_data %>% 
  ggplot(aes(x = Treatment, y = Mean_abundance, color = Saturation, group = Feature)) +
  geom_line(size=0.5, alpha=0.7) +
  scale_color_manual(values = c("SFA" = "#3969AC", "MUFA" = "#F2B701", "PUFA" = "#E73F74")) +
  labs(x = "Weeks on Western Diet",
       y = "Mean abundance (z-score)",
       color = "Saturation") +
  theme(legend.position = "right") +
  facet_wrap(~Saturation, scales = 'free')

#ggsave('../figures/lipidomics_liver_TG_saturation_change.pdf', height = 3, width = 9)
```


Read in plasma lipidomics dataset.
```{r}
plasmaData = read.csv('../processed_data/datasets/Lipidomics_plasma.csv', header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
```

Plot PCA. There are no outliers.
```{r, echo=T, fig.width=3, fig.height=2, fig.align="center", dpi=300, dev="png"}
autoplot(prcomp((plasmaData[-c(1,2)] %>% select_if(~ !any(is.na(.)))), scale. = TRUE), 
         data = plasmaData, colour = 'Treatment', 
         size = 0.1, label = TRUE, label.size = 3, frame = TRUE, frame.type = 'norm', 
         frame.level = 0.95) + 
         scale_color_manual(values = c('black', '#3b9bb3', '#75af4f', '#e39225', '#c13b41')) +
         scale_fill_manual(values = c('black', '#3b9bb3', '#75af4f', '#e39225', '#c13b41'))
```
PCA plot of plasma lipidomics data
```{r, echo=T, fig.width=3, fig.height=2, fig.align="center", dpi=300, dev="png"}
autoplot(prcomp((plasmaData[-c(1,2)] %>% select_if(~ !any(is.na(.)))), scale. = TRUE), data = plasmaData, colour = 'Treatment', size = 2.5) + scale_color_manual(values = c('black', '#3b9bb3', '#75af4f', '#e39225', '#c13b41'))

#ggsave('../figures/lipidomics_plasma_pca.pdf', height = 3.5, width = 5)
```
Bubble plot showing relative fold-change in plasma lipid abundance.
```{r, echo=T, fig.width=7, fig.height=3, fig.align="center", dpi=300, dev="png"}
colors <- c("#A00000", "#00C000", "#5757F9", "#FF6000", "#0000C0", "#C0C000", "#D7B0B0",
            "#9F044D", "#077E97", "#606060", "#fcfc81")

bubdata = makeBubbleData(plasmaData) %>% filter(!Feat %in% c("PE", "PE_O", "PE_P"))
bubp = getBubbleSignificance(bubdata)
makeBubblePlot(bubp, abundance=FALSE, colors=colors, bubble_range=c(3,10))

#ggsave('../figures/lipidomics_plasma_bubble_plot.pdf', height = 5, width = 15)
#write.csv(bubp, '../processed_data/misc/lipidomics_plasma_bubble_stats.csv')
```
Plot abundance of plasma CE species.
```{r, echo=T, fig.width=3, fig.height=7, fig.align="center", dpi=300, dev="png"}
abundant_CE = plasmaData %>% select(grep("CE", colnames(.))) %>%
  colMeans() %>% as.data.frame() %>%
  filter(. > 50)

scaled_plasmaData = cbind(plasmaData[,c(1,2)], 
                    scale(plasmaData %>% select(rownames(abundant_CE)))) # mean 0, sd 1

CE_data = scaled_plasmaData %>%
  gather(Feature, Value, 3:ncol(.))

order = c(unique((CE_data %>% filter(Treatment == 'Chow8'))$ID),
          unique((CE_data %>% filter(Treatment == 'WD2'))$ID),
          unique((CE_data %>% filter(Treatment == 'WD4'))$ID),
          unique((CE_data %>% filter(Treatment == 'WD6'))$ID),
          unique((CE_data %>% filter(Treatment == 'WD8'))$ID))

CE_data %>% ggplot(aes(x = Feature, y = as.factor(ID))) + 
  geom_tile(aes(fill = Value), alpha = 1) + 
  scale_fill_distiller(palette='RdYlBu') +
  scale_y_discrete(limits = rev(factor(order))) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        axis.title.x = element_blank(), axis.title.y = element_blank())

#ggsave('../figures/lipidomics_plasma_CE_heatmap.pdf', height = 14, width = 9)
```
Abundance of individual plasma TG species split by SFA, MUFA, and PUFA.
```{r, echo=T, fig.width=4.5, fig.height=1.5, fig.align="center", dpi=300, dev="png"}
scaled_plasmaData = cbind(plasmaData[,c(1,2)], scale(plasmaData[,-c(1,2)])) # mean 0, sd 1

mean_plasma_TG_data = scaled_plasmaData %>%
  select_if(~ !any(is.na(.))) %>%
  gather(Feature, Value, 3:ncol(.)) %>%
  separate(col = 'Feature', into = c("Class", "Chain"), sep = "\\.", 
          remove = FALSE, extra = 'drop') %>%
  filter(Class == 'TG') %>%
  group_by(Treatment, Feature, Chain) %>%
  mutate(Mean_abundance = mean(Value)) %>%
  select(Treatment, Feature, Chain, Mean_abundance) %>%
  distinct() %>%
  mutate(Saturation = case_when(
    str_extract(Feature, "TG\\.\\d+\\.\\d+") %>% 
      str_extract("\\.\\d+$") %>%
      str_replace("\\.", "") %>%
      as.numeric() == 0 ~ "SFA",
    str_extract(Feature, "TG\\.\\d+\\.\\d+") %>% 
      str_extract("\\.\\d+$") %>%
      str_replace("\\.", "") %>%
      as.numeric() == 1 ~ "MUFA",
    str_extract(Feature, "TG\\.\\d+\\.\\d+") %>% 
      str_extract("\\.\\d+$") %>%
      str_replace("\\.", "") %>%
      as.numeric() > 1 ~ "PUFA",
    TRUE ~ NA_character_
  )) %>% mutate(Treatment = as.factor(Treatment)) %>% mutate(
    Treatment = fct_recode(
      Treatment,
      "0" = "Chow8",
      "2" = "WD2",
      "4" = "WD4",
      "6" = "WD6",
      "8" = "WD8"
  )
)

mean_plasma_TG_data %>% 
  ggplot(aes(x = Treatment, y = Mean_abundance, color = Saturation, group = Feature)) +
  geom_line(size=0.5, alpha=0.7) +
  scale_color_manual(values = c("SFA" = "#3969AC", "MUFA" = "#F2B701", "PUFA" = "#E73F74")) +
  labs(x = "Weeks on Western Diet",
       y = "Mean abundance (z-score)",
       color = "Saturation",
       title = "Plasma TG") +
  theme(legend.position = "right") +
  facet_wrap(~Saturation, scales = 'free')

#ggsave('../figures/lipidomics_plasma_TG_saturation_change.pdf', height = 3, width = 9)
```
Spearman correlation between hepatic and plasma lipids.
```{r, echo=T, fig.width=2.5, fig.height=1.5, fig.align="center", dpi=300, dev="png"}
# correlate total lipid classes between liver and plasma
totalLipidbyClass_liver = liverData %>%
  gather(Feature, Value, 3:ncol(.)) %>%
  separate(col = 'Feature', into = c("Class"), sep = "\\.", remove = FALSE, extra = 'drop') %>%
  mutate(RefinedClass = ifelse(grepl("PE[.]O", Feature), "PE_O",
                        ifelse(grepl("PE[.]P", Feature), "PE_P", Class))) %>%
  select(-Class) %>%
  group_by(ID, RefinedClass) %>%
  drop_na() %>%
  mutate(totalLipid = sum(Value)) %>%
  ungroup() %>%
  select(ID, Treatment, RefinedClass, totalLipid) %>%
  distinct() %>%
  group_by(RefinedClass) %>%
  mutate(n_groups = length(unique(Treatment))) %>%
  filter(n_groups == 5) %>%
  ungroup() %>%
  select(-n_groups)

totalLipidbyClass_plasma = plasmaData %>%
  gather(Feature, Value, 3:ncol(.)) %>%
  separate(col = 'Feature', into = c("Class"), sep = "\\.", remove = FALSE, extra = 'drop') %>%
  mutate(RefinedClass = ifelse(grepl("PE[.]O", Feature), "PE_O",
                        ifelse(grepl("PE[.]P", Feature), "PE_P", Class))) %>%
  select(-Class) %>%
  group_by(ID, RefinedClass) %>%
  drop_na() %>%
  mutate(totalLipid = sum(Value)) %>%
  ungroup() %>%
  select(ID, Treatment, RefinedClass, totalLipid) %>%
  distinct() %>%
  group_by(RefinedClass) %>%
  mutate(n_groups = length(unique(Treatment))) %>%
  filter(n_groups == 5) %>%
  ungroup() %>%
  select(-n_groups)
 
lipids_in_common = intersect(totalLipidbyClass_liver$RefinedClass, totalLipidbyClass_plasma$RefinedClass)
 
df = data.frame("LipidClass" = character(), "Spearman_corr" = numeric(), "p_val" = numeric())
for (lipid1 in lipids_in_common) {
  lipid_in_liver = totalLipidbyClass_liver %>% filter(RefinedClass == lipid1)
  for (lipid2 in lipids_in_common) {
    lipid_in_plasma = totalLipidbyClass_plasma %>% filter(RefinedClass == lipid2)
    combined = left_join(lipid_in_liver, lipid_in_plasma, by = 'ID')
    corr = cor.test(combined$totalLipid.x, 
                    combined$totalLipid.y, 
                    method=c("pearson", "kendall", "spearman"))$estimate
    p = cor.test(combined$totalLipid.x, 
                 combined$totalLipid.y, 
                 method=c("pearson", "kendall", "spearman"))$p.value
    df = rbind(df, data.frame("Lipid_liver" = lipid1,
                              "Lipid_plasma" = lipid2,
                              "Spearman_corr" = corr,
                              "p_val" = p))
  }
}

# save dataframe
write.csv(df, '../processed_data/misc/liver_plasma_lipid_class_pairwise_corr.csv', row.names=FALSE)

df %>%
  ggplot(aes(x=Lipid_liver, y=Lipid_plasma)) +
  geom_tile(aes(fill=Spearman_corr), alpha=1, color='black', size=0.1) +
  scale_fill_gradient2(low = "#0000FF", mid = "#FFFFFF", high ="#FF0000", midpoint=0) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#ggsave('../figures/lipidomics_plasma_liver_total_corr_heatmap.pdf', height = 4, width = 6)
```

```{r}
# rmarkdown::render("lipidomics.Rmd", output_format="github_document", output_dir="../markdowns")
```
